A Classical Density-Functional Theory for Describing Water Interfaces

### 2011-09-26 paper/project.tex
Project text and bib edits. Tests skipped!

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
{We develop a classical density functional theory for water by combining the fundamental-measure theory functional--which describes hard sphere fluids--with attractive interactions based on the Statistical Associating Fluid Theory (SAFT). Using this approximation, we are able to describe the density, free energy, and
entropy of water, as well as the number of hydrogen bonds formed near liquid-vapor and liquid-solid interfaces. These interfaces include water confined to a hard-walled slit, a single hydrophobic rod immersed in water, and two hydrophobic rods 
at various separations. We found \needsworknow{conclusion-like stuff about two rods, vapor-liquid transition between rods, constant force of twice surface tension, more?}.
This functional provides a foundation for future work predicting the interactions of chemicals in aqueous solution on an
atomic level.}

@@

Classical density-functional theory (DFT) provides an efficient way to evaluate
the free energy using the probability distribution of the fluid. The free energy 
is a functional
of the density function and is independent of any external
potentials\cite{ebner1976}. The most common use of DFT is calculating ground
state electron densities in quantum mechanical systems. The Hohenberg-Kohn
theorem\cite{hohenberg1964inhomogeneous} allows us to write the ground state
energy in terms of a universal functional of the electron density and a separate
term including the external potential. Mermin\cite{mermin1965thermal} gives a
theorem extending that of Hohenberg and Kohn to nonzero temperatures:
\begin{equation}
  A(T) = \underset{n(r)}{min}\left\{ F[n(r),T] + \int V_{ext}(r) n(r)
d^3r\right\}
\end{equation}
This is the basis for classical DFT, where the relevant density is not the
electron density,
but the density of molecules. This can be used with a variety of 
external potentials, from the simple interaction with a 
hydrophobic wall to the more complex, stronger interaction with hydrophilic
solutes.

@@

For reference, we plot the coexistence curve (see
Figure~\ref{fig:temperature-vs-density}) and the predicted vapor pressure (see
Figure~\ref{fig:equation-of-state}) against experimental values.  These
match those calculated by Clark \emph{et al}. The functional provides a good description of
the vapor pressure, which has an error of less than 
1\% of the experimental data. The coexistence curve is calculated by 
\fixme{explanation of coexistence curve calculation?}. Near the critical temperature,
the agreement is off, and the theoretical prediction  of the temperature is quite a bit higher than
experiment. To find the densities, we use the
``common tangent'' rule, as illustrated in Figure~\ref{fig:homogeneous}.
\needsworknow{Expand the previous sentence - common tangent rule}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/density-vs-radius}
\end{center}
\caption{ Density profiles for 
a single hydrophobic rod with a radius of 0.5~nm using several different 
lattice spacing resolutions. We used a high resolution of 0.05~bohrs ($\approx$
0.003~nm), 
medium 0.2~bohrs ($\approx$ 0.001~nm), and low 0.5~bohrs ($\approx$ 0.03~nm).}
\label{fig:densityresolution}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
{We develop a classical density functional theory for water by combining the
fundamental-measure theory functional--which describes hard sphere fluids--with
attractive interactions based on the Statistical Associating Fluid Theory
(SAFT). Using this approximation, we are able to describe the vapor pressure, 
various densities, surface tension, enthalpy and
entropy of water, as well as the number of hydrogen bonds formed near
liquid-vapor and liquid-solid interfaces. These interfaces include water
confined to a hard-walled slit, a single hydrophobic rod immersed in water, and
two hydrophobic rods 
at various separations. We found \needsworknow{conclusion-like stuff about two
rods, vapor-liquid transition between rods, constant force of twice surface
tension, more?}.
This functional provides a foundation for future work predicting the
interactions of chemicals in aqueous solution on an
atomic level.}

@@

Classical density-functional theory (DFT) provides an efficient way to evaluate
the free energy using the probability distribution of the fluid. The free energy 
is a functional
of the density function and is independent of any external
potentials\cite{ebner1976density}. The most common use of DFT is calculating ground
state electron densities in quantum mechanical systems. The Hohenberg-Kohn
theorem\cite{hohenberg1964inhomogeneous} allows us to write the ground state
energy in terms of a universal functional of the electron density and a separate
term including the external potential. Mermin\cite{mermin1965thermal} gives a
theorem extending that of Hohenberg and Kohn to nonzero temperatures:
\begin{equation}
  A(T) = \underset{n(r)}{min}\left\{ F[n(r),T] + \int V_{ext}(r) n(r)
d^3r\right\}
\end{equation}
This is the basis for classical DFT, where the relevant density is not the
electron density,
but the density of molecules. This can be used with a variety of 
external potentials, from the simple interaction with a 
hydrophobic wall to the more complex, stronger interaction with hydrophilic
solutes.

@@

For reference, we plot the coexistence curve (see
Figure~\ref{fig:temperature-vs-density}) and the predicted vapor pressure (see
Figure~\ref{fig:equation-of-state}) against experimental values.  These
match those calculated by Clark \emph{et al}. The functional provides a good description of
the vapor pressure, which has an error of less than 
1\% of the experimental data. Near the critical temperature,
the agreement is off, and the theoretical prediction  of the temperature is quite a bit higher than
experiment. To find the densities for the coexistence curve, we use the
``common tangent'' rule. The two densities 
will be on the same tangent line to the free energy as illustrated in the top of Figure~\ref{fig:homogeneous}.
As the critical point approaches, the vapor and liquid coexistence densities get closer and closer together 
(see the bottom of Figure~\ref{fig:homogeneous}. 
\needsworknow{Expand the previous common tangent rule stuff?}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/density-vs-radius}
\end{center}
\caption{ Density profiles for 
a single hydrophobic rod with a radius of 0.5~nm using several different 
lattice spacing resolutions. We used a high resolution of 0.05~bohrs ($\approx$~0.003~nm), 
medium 0.2~bohrs ($\approx$~0.001~nm), and low 0.5~bohrs ($\approx$~0.03~nm).}
\label{fig:densityresolution}
\end{figure}
